---
title: "BellaBeat Analysis"
author: "Adam Stacey"
date: "7/19/2021"
output: html_document
---


## Getting Started

We've been tasked by BellaBeat to identify current smart device trends.  We'll use
these insights to identify strategies BellaBeat can incorporate to increase their
market share.  We've been provided a public dataset to work with.

We have 31 days worth of data for 33 Fitbit users.  This data has been made 
public and is available on [kaggle](https://www.kaggle.com/arashnic/fitbit).  
33 users isn't a lot, it would be better if the sample size was larger.  However, 
the data produced by these smart devices is *quite* extensive.

The data is broken down into several tables.  The first tables are of a manageable 
size and could be analyzed in a spreadsheet program.  However, the other tables 
are very long as they contain per hour, per minute, and per second data.  So I'll 
use R as my main tool for this analysis.  

I'll start with loading the all powerful tidyverse.  I'll load lubridate too for
help with dates.
  
```{r, message = FALSE}
library(tidyverse)
library(lubridate)
```
  
Let's load the tables that track daily activity into tibbles.
  
```{r}
daily_activity_tibble <- as_tibble(read.csv('../input/fitbit/Fitabase Data 4.12.16-5.12.16/dailyActivity_merged.csv'))
glimpse(daily_activity_tibble)

```

This table shows up to 31 days of activity for the 33 Fitbit users.  It includes
total steps, distance, calories burned, along with distance/minutes in different 
cardiovascular zones (sedentary, light, moderate, and very active).

I'll load the total daily sleep table into a tibble too.

```{r}
daily_sleep_tibble <- as_tibble(read.csv('../input/fitbit/Fitabase Data 4.12.16-5.12.16/sleepDay_merged.csv'))
glimpse(daily_sleep_tibble)
```

This table shows the total minutes asleep, total minutes in bed and the number of
sleep events for each user daily.

The last daily table is a weight log.  PROCESSING...  

```{r}
daily_weightlog_tibble <- as_tibble(read.csv('../input/fitbit/Fitabase Data 4.12.16-5.12.16/weightLogInfo_merged.csv'))
glimpse(daily_weightlog_tibble)
```

This short table shows a users weight and BMI on a daily basis.  Most values are 
entered manually.  


### Cleaning

Let's remove any duplicate rows.

```{r}
daily_activity_tibble <- daily_activity_tibble %>% distinct()
daily_sleep_tibble <- daily_sleep_tibble %>% distinct()
daily_weightlog_tibble <- daily_weightlog_tibble %>% distinct()
```

Let's add columns to express the 'minutes asleep' and 'minutes in bed' columns 
in hours since we usually think of our sleep time in hours. 

```{r}
daily_sleep_tibble <- daily_sleep_tibble %>% 
  mutate(TotalHoursAsleep = TotalMinutesAsleep / 60,
         TotalHoursInBed = TotalTimeInBed / 60)
glimpse(daily_sleep_tibble)
```

## Daily Activity Insights

One interesting thing Id like to find out is: how many days during the month did 
each of these users wear their smart device?

```{r}
days_used <- daily_activity_tibble %>% count(Id)
days_used %>% 
  ggplot(aes(x = as.factor(Id), y = n)) +
  geom_col(fill = 'grey', color = 'black', width = 0.5) +
  labs(title = 'Number of Days Used', subtitle = '04/12/16 through 05/12/16',
       x = 'Users', y = 'Number of Days') +
  theme(axis.text.x = element_blank())
```

### Analysis

As you can see, most users did use their device for 30 or 31 days total.  However,
you can see that some used their device significantly less.  Why is this?  More
data would be needed for a definitive answer, but perhaps after charging the device,
some users simply forgot to put it back on for a day or more.  Some users may only
wear their device on days they work out.  Or they may have multiple devices that
they rotate through.

## Sleep Tracking Insights

Let's ask the same for daily sleep activity:  how many nights during the month 
did each of these users wear their smart device?

```{r}
nights_used <- daily_sleep_tibble %>% count(Id)
nights_used %>% 
  ggplot(aes(x = as.factor(Id), y = n)) +
  geom_col(fill = 'grey', color = 'black', width = 0.5) +
  labs(title = 'Number of Nights Used', subtitle = '04/12/16 through 05/12/16',
       x = 'Users', y = 'Number of Days') +
  theme(axis.text.x = element_blank())
```

Here we see a wildly different trend than in our tracking of daytime use.  12 of 
the users did use their device to track sleep for most or all of the time.  12 
users had significantly less use at night, with most being below 10 nights in the 
month.  And please note that 9 users didn't use the device at night at all, so 
they do not even appear on the chart.

Another interesting question to ask is:  how long total do users stay in bed 
versus their actual sleep time?  We'll take the average for each user and display
it in a bar chart.

```{r}
daily_sleep_tibble <- daily_sleep_tibble %>% 
  mutate(Extra_Time_Bed = TotalTimeInBed - TotalMinutesAsleep) %>% 
  group_by(Id) %>% 
  summarize(mean = mean(Extra_Time_Bed))

ggplot(data = daily_sleep_tibble) +
  geom_col(mapping = aes(x = as.factor(Id), y = mean), fill = 'grey', 
                         color = 'black', width = 0.5) +
  labs(title = 'Extra Time Spent in Bed', subtitle = 'Average', x = 'Users',
       y = 'Extra Time in Bed (Minutes)') +
  theme(axis.text.x = element_blank(), legend.position = 'none') +
  scale_y_continuous(breaks=seq(0,350,50))
```

As you can see, most users spend less than 50 extra minutes in bed on average.  
It is unknown to me if this includes time before falling asleep.  Presumably the 
device measure heart rate to determine that a user is still at rest in bed but no
longer asleep.  Two users spent significantly more time in bed on average.  Were 
they sick?

### Analysis

Many users are not using their device at night regularly if at all.  Again, a 
probable cause is that they charge it at night.  Or perhaps they find the device 
uncomfortable while sleeping.  Maybe many don't consider their sleep data to be 
very important.  A survey would be helpful in determining the cause.  

## Weight Log Insights

There are only 67 entries in this table compared to a possible total of 1,023
(if all 33 users entered a value all 31 days).  Most entries are done manually, 
presumably meaning the user measured themselves on a scale and put the value in 
the app.  A few were not, probably meaning they used a smart scale that is also
connected to the app.  There are only two entries for fat (percentage of body 
fat I believe).

### Analysis

People aren't using this feature much.  There could be many causes for this.  It
could well just be psychological.  I personally don't think I would want to 
measure my weight every day.  It could also be a matter of being too much work.  
Most probably don't have an integrated smart scale, and so weighing yourself and 
manually entering it in the app each day may seem tedious.  As for the fat column, 
how is it being measured?  More than just weight and height must be involved or
the column should always be automatically filled out.  Survey data would be 
helpful to find out why people don't use this feature much. 

## Deeper dive

There is plenty of data left.  What should we look at?  I'm curious about METs.
I didn't know what they were but looked it up.  MET stands for metabolic 
equivalent. 1 MET is your resting metabolic rate. An activity that is 4 METs 
requires your body to consume 4 times the oxygen you would when at rest. That 
activity obviously requires more energy and leads to more calories burned.   

I'll load the MET table into a tibble and clear any duplicate entries.  We also
need to cast the ActivityMinute field as a date instead of a string.

```{r}
minute_MET_tibble <- as_tibble(read.csv('../input/fitbit/Fitabase Data 4.12.16-5.12.16/minuteMETsNarrow_merged.csv'))
minute_MET_tibble <- minute_MET_tibble %>% distinct()
glimpse(minute_MET_tibble)

minute_MET_tibble$ActivityMinute <-mdy_hms(minute_MET_tibble$ActivityMinute) 
glimpse(minute_MET_tibble)
```

You may have noticed in the glimpse that the METs are very high.  I'd expect to
see a lot of 1s, especially at night but in fact, in all 1.3 million rows there 
are only a handful of values below 10. And  there are a couple of thousand rows 
with values over 100. To me, all the values seem to be 10 times what they should 
be.  I haven't been able to find an answer as to why this is.

Let's pick a user and see how their MET's fluctuate over the course of a couple 
hours.

```{r}
filtered_minute_MET_tibble <- minute_MET_tibble %>% 
  filter(Id == 5553957443 & ActivityMinute >= '2016-04-15 15:00:00' & 
           ActivityMinute < '2016-04-15 17:00:00')

ggplot(data = filtered_minute_MET_tibble) +
  geom_line(mapping = aes(x = ActivityMinute, y = METs)) +
  labs(title = 'MET Fluctuation', subtitle = "Between 3 and 5 PM", x = 'Minute') +
  scale_y_continuous(breaks=seq(0,45,15))
```

You can see the fluctuation.  Interesting, but not particularly useful in terms 
of smart device trends.

I'm going to pull up the hourly step table now:

```{r}
hourly_step_tibble <- as_tibble(read.csv('../input/fitbit/Fitabase Data 4.12.16-5.12.16/hourlySteps_merged.csv'))
hourly_step_tibble <- hourly_step_tibble %>% distinct()
hourly_step_tibble$ActivityHour <-mdy_hms(hourly_step_tibble$ActivityHour) 
glimpse(hourly_step_tibble)
```

I also cleared duplicate entries and cast the ActivityHour column as a date 
instead of a string.  Now let's find the most active hour for each user.

```{r}
most_step_hour <- hourly_step_tibble %>% 
  group_by(Id) %>% 
  filter(StepTotal == max(StepTotal)) 

most_step_hour$ActivityHour <- hour(most_step_hour$ActivityHour)

glimpse(most_step_hour)

ggplot(data = most_step_hour) +
  geom_bar(mapping = aes(x = ActivityHour), fill = 'gray', color = 'black') +
  scale_x_continuous(breaks=seq(0,23,1)) +
  labs(title = 'Hour with the Highest Steps', x = 'Hour, military')

```

As you can see, the hour with the most steps is 9 AM for five users.  8 AM has 
the second most with 4 users.  Most likely as people get up and ready for work, 
they end up walking a lot.  

## Conclusions

My goal was to identify trends for smart device users.  The sample size I had to 
work with is small but the amount of data generated is staggering.  I don't think
most of it will help identify trends that lead to actionable insights, but 
there are some things worth sharing.  

As we saw earlier, most people do use their device everyday or almost everyday.
Some use their device less.  My hypothesis is that people forget to put the watch
back on after charging it.  As more and more people return to the office, it 
becomes more likely that once a person forgets to put the device on, it will 
remain off for the whole day.  Survey data would be needed to confirm but if 
correct, a longer battery life would probably result in people using their device 
more.  The more they use it, the more dependent on it they become, making them
come back for the newer, better model when it is released!

When it comes to night-time use, people tend to use their device much less 
frequently.  Probable causes are that the device is charging at night, or is 
uncomfortable.  Longer battery life could increase their use.  A smaller device
may make it more comfortable to wear at night.  But also, just more education
regarding sleep and how important it is and how it affects your health may 
prompt users to use their device at night. This could take the form of short, 
informative pop-ups when people open our app.  



